---
title: "Introduction to postGIStools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{postGIStools-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
# Don't run this vignette on CRAN
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(purl = NOT_CRAN, eval = NOT_CRAN)
knitr::opts_chunk$set(warning=FALSE, message=FALSE, collapse=TRUE, comment = "#>")
```

The postGIStools package extends the standard R / PostgreSQL interface (as implemented in RPostgreSQL) to provide support for two popular PostgreSQL extensions: _PostGIS_ (spatial data) and _hstore_ (key/value pairs). 


## Reading PostGIS and hstore data into R

We demonstrate the postGIStools functions using a test database hosted on Heroku. It contains a single table _country_ with the following fields:

| name      | type   | comments |
|-----------|--------|-------------------------------------|
| name      | text   | country name |
| iso2      | text   | ISO two-letter code (primary key) |
| capital   | text   | |
| population| integer | |
| translations| hstore | key/value pairs where key is language (e.g. "es", "fr") |
| geom      | geometry | country polygons |


The data originates from the [REST Countries API](https://restcountries.eu/), whereas the country geometries are from the _wrld\_simpl_ map included in maptools R package.

_Note: Currently, the test database cannot be accessed from a Windows or Mac OS X environment, due the lack of SSL support (which is required by Heroku) in those versions of RPostgreSQL._

To read data from PostgreSQL into R, postGIStools provides the `get_postgis_query` function. Like the `dbGetQuery` function in PostgreSQL, it requires a connection object and a SQL statement, which in this case must be a SELECT statement. In addition, the user may indentify a geometry and/or hstore field by name. Note that since the geometry and hstore field names must be found in the query statement, wildcards ("SELECT *") cannot be used.

```{r get_postgis_query}
library(RPostgreSQL)
library(postGIStools)

con <- dbConnect(PostgreSQL(), dbname = "d2u06to89nuqei", user = "mzcwtmyzmgalae",
                 host = "ec2-107-22-246-250.compute-1.amazonaws.com",
                 password = "UTv2BuwJUPuruhDqJthcngyyvO")

countries <- get_postgis_query(con, "SELECT name, iso2, capital, population,
                               translations, geom FROM country 
                               WHERE population > 1000000",
                               geom_name = "geom", hstore_name = "translations")

class(countries)
```

When a geometry column is specified, the query output is a spatial data frame type from the sp package. The hstore column is converted to a list-column in R, where each "cell" is a named list.

```{r data_str}
str(countries@data[1:2,])
```


## Working with hstore columns

To interact with hstore columns imported into R, postGIStools defines the `%->%` operator, which is analogous to `->` in PostgreSQL. Specifically, `hstore %->% "key"` extracts the value in each cell of the hstore corresponding to the given key, or `NA` when the key is absent from a given cell.

```{r hstore_select}
head(countries$translations %->% "es")
```

The operator is also compatible with single bracket subsetting of the hstore.
```{r hstore_subset}
countries$translations[5:7] %->% "fr"
```

The assignment version of `%->%` operates similarly, with the option of deleting keys by assigning them to `NULL`.
```{r hstore_assign}
countries$translations[2] %->% "nl" <- "Albanië"
countries$translations[3] %->% "fr" <- NULL
countries$translations[2:3]
```

The `new_hstore` function creates a blank hstore of a given length, which is just an empty list of lists. It is most useful when assigned to a data frame column (e.g. `df$hs <- new_hstore(3)`) that can be then populated with `%->%` and written back to a PostgreSQL database.


## Inserting and updating PostgreSQL tables from R

The two write methods `postgis_insert` and `postgis_update` wrap around their namesake SQL commands, while also converting R spatial objects and list-columns back into the geometry and hstore data types, respectively.

To demonstrate these functions, we create a new temporary table in the database.
```{r create_tmp_table, results = "hide"}
dbSendQuery(con, paste("CREATE TEMP TABLE cty_tmp (name text,", 
                       "iso2 text PRIMARY KEY, capital text,",
                       "translations hstore, geom geometry)"))
```

Calls to `postgis_insert` must specify the connection, data frame and table name. By default, all data frame columns are inserted, but a subset of columns can be specified as `write_cols`. In both cases, the names of inserted columns must have a match in the target table.
```{r postgis_insert}
postgis_insert(con, countries[1:10,], "cty_tmp",
               write_cols = c("name", "iso2", "translations"),
               geom_name = "geom", hstore_name = "translations")

# Reimport to check
cty_tmp <- get_postgis_query(con, paste("SELECT name, iso2, capital,",
                                        "geom, translations FROM cty_tmp"),
                             geom_name = "geom", hstore_name = "translations")
head(cty_tmp@data)
```

We next update the records in _cty\_tmp_ to include the _capital_ field. The syntax of `postgis_update` is similar to `postgis_insert`, except that we must specify both `id_cols`, the column(s) identifying the records to update, as well as `update_cols`, the column(s) to be updated. (The underlying PostgreSQL operation is of the format _UPDATE... SET ... FROM..._.) Neither the geometry nor the hstore can be used as `id_cols`. Note that since the input data frame `countries[1:10,]` includes spatial and list-column data, we need to specify `geom_name` and `hstore_name`, even if those columns are not needed for the update operation.

```{r postgis_update}
postgis_update(con, countries[1:10,], "cty_tmp", id_cols = "iso2", 
               update_cols = "capital", geom_name = "geom", 
               hstore_name = "translations")

cty_tmp <- get_postgis_query(con, paste("SELECT name, iso2, capital,",
                                        "geom, translations FROM cty_tmp"),
                             geom_name = "geom", hstore_name = "translations")
head(cty_tmp@data)
```

By default, hstore columns are updated by concatenation: keys present in the input data frame but not the original table are added to the hstore, keys present in both the data frame and table have their associated values updated, but keys absent from the input data frame are _not_ deleted from the table. This can be changed by setting `hstore_concat = FALSE`, in which case whole hstore cells are replaced with corresponding ones in the input data frame.

```{r update_hstore}
countries$translations[2] %->% "nl" <- NULL
countries$translations[3] %->% "fr" <- "Algérie"
 
postgis_update(con, countries[1:10,], "cty_tmp", id_cols = "iso2", 
               update_cols = "translations", geom_name = "geom", 
               hstore_name = "translations")

cty_tmp <- get_postgis_query(con, paste("SELECT name, iso2, capital,",
                                        "geom, translations FROM cty_tmp"),
                             geom_name = "geom", hstore_name = "translations")
cty_tmp@data[cty_tmp$iso2 %in% c("AL", "DZ"), ]

# Key deletion not reflected in database unless hstore_concat = FALSE
postgis_update(con, countries[1:10,], "cty_tmp", id_cols = "iso2", 
               update_cols = "translations", geom_name = "geom", 
               hstore_name = "translations", hstore_concat = FALSE)

cty_tmp <- get_postgis_query(con, paste("SELECT name, iso2, capital,",
                                        "geom, translations FROM cty_tmp"),
                             geom_name = "geom", hstore_name = "translations")
cty_tmp@data[cty_tmp$iso2 %in% c("AL", "DZ"), ]
```
